\chapter{Circuit Analysis}


\section{Electric Current}

The \textbf{electric current} is defined as the rate at which \textbf{charges}
$Q$ pass through a point in a circuit. In differential form, it is expressed as
a function of time:
\begin{equation}
  \boxed{I(t)=\diff Qt} %= \frac QV\diff Vt = (ne)(Av_d)
  \label{eq:current}
\end{equation}
The SI unit for electric current is an \emph{amp\`{e}re} (\si\ampere), and is
one of the seven base SI unit. (The origin of the definition of an amp\`{e}re
of current was discussed in Chapter~\ref{chapter:magnetism1}.) The SI unit for
electrical charge, a \emph{coulomb}, is therefore defined as the amount of
charge that \SI1{\ampere} of current passes through a circuit in \SI1\second.


\subsection{DC vs.\ AC}
In a \textbf{direct current} (\textbf{DC}), the flow of charges is always in
the same direction, but the current itself does not have to be constant in
time.
\begin{figure}[ht]
  \centering
  \begin{subfigure}{.3\linewidth}
    \centering
    \begin{tikzpicture}[scale=.8]
      \draw[function] (0,3)--(4,3);
      \draw[axes] (0,0)--(4,0) node[right]{$t$};
      \draw[axes] (0,0)--(0,4) node[above]{$I$};
    \end{tikzpicture}
  \end{subfigure}
  \begin{subfigure}{.3\linewidth}
    \centering
    \begin{tikzpicture}[scale=.8]
      \draw[smooth,samples=20,domain=0:3.5,function]
      plot(\x,{3*(exp(-.8*\x))});
      \draw[axes] (0,0)--(4,0) node[right]{$t$};
      \draw[axes] (0,0)--(0,4) node[above]{$I$};
    \end{tikzpicture}
  \end{subfigure}
  \begin{subfigure}{.3\linewidth}
    \centering
    \begin{tikzpicture}[scale=.8]
      \draw[smooth,samples=20,domain=0:3.5,function] plot(\x,{sin(120*\x)+2});
      \draw[axes] (0,0)--(4,0) node[right]{$t$};
      \draw[axes] (0,0)--(0,4) node[above]{$I$};
    \end{tikzpicture}
  \end{subfigure}
\end{figure}

In an \textbf{alternating current} (AC), the flow of charges changes
direction, usually as a sinusoidal function of time.
\begin{figure}[ht]
  \centering
  \begin{subfigure}{.3\linewidth}
    \centering  
    \begin{tikzpicture}[scale=.8]
      \draw[axes] (0,0)--(5,0) node[right]{$t$};
      \draw[axes] (0,-2)--(0,2) node[above]{$I$};
      \draw[smooth,samples=50,domain=0:4,function] plot(\x,{1.5*sin(150*\x)});
    \end{tikzpicture}
  \end{subfigure}
  \begin{subfigure}{.3\linewidth}
    \centering
    \begin{tikzpicture}[scale=.8]
      \draw[axes] (0,0)--(5,0) node[right]{$t$};
      \draw[axes] (0,-2)--(0,2) node[above]{$V$};
      \draw[smooth,samples=40,domain=0:4,function] plot(\x,{1.5*sin(150*\x)});
    \end{tikzpicture}
  \end{subfigure}
\end{figure}
The power outlet in North America are all AC, with a maximum voltage of
\SI{120}\volt, and a frequency of \SI{60}\hertz. AC current are important
in the power generation.



\subsection{Quantifying Electric Current}
%The amount of current that pases through a circuit depends on several factors,
%namely:
%\begin{itemize}[leftmargin=15pt]
%\item The charge that each ``charge carrier'' carries. (We already know that
%  in a metal conductor, the charge carriers are electrons, with an
%  elementary charge of $e=\SI{1.602e-19}\coulomb$.)
%\item How many charges are moving through the circuit. This then depends on
%  the density of the charge carriers inside the conducting material, as well as
%  the cross sectional area $A$ of the conductor.
%\item The speed $v_d$ in which the charges moving through the circuit.
%\end{itemize}
We can further expand Eq.~\ref{eq:current} to examine the exact quantities that
the current depends on:
\begin{equation}
  I(t)=\diff Qt=\left[\frac QV\right]\diff Vt
\end{equation}
where
\begin{itemize}
\item $Q/V$ is charge density, which is the amount of charge per volume,
  as discussed in Section~\ref{section:charge-density}. This is the charge of
  each charge carrier ($e$)\footnote{We have established that in a metal
  conductor, the charge carriers are electrons, with a charge of
  $e=\SI{1.602e-19}\coulomb$.}, multiplied by the \emph{charge carrier density}
  ($n$).
\item $\diff Vt$ is the rate the at which the volume of charges moves through
  the conductor. This is just the ``volume flux'' that was discussed in
  Chapter~\ref{chapter:gauss}, and it is the speed $v_d$ at which the charges
  are moving, called the \textbf{drift velocity}, multiplied by the
  cross-sectional area of the coductor $A$.
\end{itemize}
%  is just
%  the \textbf{charge carrier density} (number of charge carriers per volume)
%  $n$ times the \textbf{elementary charge} $e$
%\item $\diff V/t$ is the

%, give by the wire's cross-section area $A$ times the
%  \textbf{drift velocity} $v_d$ of the charge carrier
%\end{itemize}
Collecting the terms, we arrive at the equation:
\begin{equation}
  \boxed{
    I(t)=neAv_d
  }
\end{equation}

Calculating the charge carrier density ($n$) in a \emph{metal} conductor
involves a few steps based on the physical information about the metal:
\begin{enumerate}
\item Divide the metal's density $\rho$ by its molar mass $M$ to find the
  \emph{number of moles of atoms per \si{\metre^3}}
\item Multiply by Avogadro's number $N_A=\num{6.0221e23}$ to find
  \emph{number of atoms per \si{\metre\cubed}}
\item Multiply by the number of free electrons per atom $k$ for that particular
  metal to find \emph{the number of charge carriers per \si{\metre\cubed}}
\end{enumerate}
Collecting all the terms from the last slide, we have:
\begin{equation}
  \boxed{n=\frac{\rho kN_A}M}
\end{equation}
%\begin{center}
%  \begin{tabular}{l|c|c}
%    \rowcolor{pink}
%    \textbf{Quantity} & \textbf{Symbol} & \textbf{SI Unit} \\ \hline
%    Charge carrier density   & $n$    & \si{\per\metre\cubed} \\
%    Density of material      & $\rho$ & \si{\kilo\gram\per\metre\cubed} \\
%    Free electrons per atom  & $k$    & \\
%    Avogadro's number        & $N_A$  & \si{\per\mol}\\
%    Molar mass               & $M$    & \si{\kilo\gram\per\mol}
%  \end{tabular}
%\end{center}
For copper, $M=\SI{63.54e-3}{\kilo\gram\per\mol}$,
$\rho=\SI{8.96e3}{\kilo\gram\per\metre\cubed}$, $k=1$ and therefore
$n=\SI{8.5e28}{\per\metre\cubed}$.
%The drift velocity is in the order of
%$v\approx\SI1{\milli\metre\per\second}$.

An alternate description of the electric current is to express it in
terms of the \textbf{current density} $J$, with a unit of of \emph{amp\`{e}re
per meters squared} (\si{\ampere\per\meter\squared}).\footnote{We had
previously encountered this quantity when studying Amp\`{e}re's law.}
\begin{equation}
  \boxed{I(t)=J(t)A}
\end{equation}
It is obvious from the previous expression that the current density is the
product of the charge carrier density, elementary charge, and the drift
velocity:
\begin{equation}
  \boxed{J=nev_d}
\end{equation}



\subsection{Electric Field}
Charge carriers move inside the circuit because they are subjected

\subsection{Electric Current: Conventional vs.\ Electron Flow}
The flow of electric current assumes the flow of \emph{positive} charges. We
call this the \textbf{conventional current}:
\begin{center}
  \begin{tikzpicture}[scale=.8]
    \draw (0,0)--(10,0);
    \draw (0,1)--(10,1) node[midway,above]{$I\longrightarrow$};
    \foreach \x in {1,3,...,9}{
      \draw[thick,fill=pink!40] (\x,.5) circle (.25) node{$+$};
      \draw[vector] (\x+.25,.5)--(\x+.75,.5);
    }
  \end{tikzpicture}
\end{center}
In a conducting wire, however, negatively charged electrons flow in the
opposite direction. We call this the \textbf{electron current}:
\begin{center}
  \begin{tikzpicture}[scale=.8]
    \draw (0,0)--(10,0);
    \draw (0,1)--(10,1) node[midway,above]{$I\longrightarrow$};
    \foreach \x in {1,3,...,9}{
      \draw[thick,fill=blue!40!gray!20] (\x,.5) circle (.25) node{$-$};
      \draw[vector] (\x-.25,.5)--(\x-.75,.5);
    }
  \end{tikzpicture}
\end{center}
%Even though it is actually electron that are moving, we will continue to
%treat electric current as the flow of positive charges in the direction of
%the conventional current.




\section{Resistors}
The resistivity of a material is proportional to the electric field and
current density:

\begin{equation}
  \boxed{\bm E=\rho\bm J}
  \quad\text{or}\quad
  \boxed{\rho=\left|\frac EJ\right|}
\end{equation}
\begin{center}
  \begin{tabular}{l|c|c}
    \rowcolor{pink}
    \textbf{Quantity} & \textbf{Symbol} & \textbf{SI Unit} \\ \hline
    Electric field & $\bm E$ & \si{\newton\per\coulomb} \\
    Current density & $\bm J$ & \si{\ampere\per\metre\squared} \\
    Resistivity & $\rho$ & \si{\ohm\metre}
  \end{tabular}
\end{center}
\begin{itemize}
\item In a conductor, the electrons are free to move, and the electric
  field tend to be weak, and the resistivity is low.
\item In an insulator, electrons cannot move easily, therefore the electric
  field are generally strong, and the resistivity is high.
\end{itemize}




\section{Resistance of a Conductor}
The resistance of a conductor is proportional to the resistivity $\rho$ and
its length $L$, and inversely proportional to the cross-sectional area $A$:
\begin{equation}
  \boxed{
    R=\int\dl R=\rho\int_0^L\frac{\dl x}{A(x)}
  }
\end{equation}
%\begin{center}
%  \begin{tabular}{l|c|c}
%    \rowcolor{pink}
%    \textbf{Quantity} & \textbf{Symbol} & \textbf{SI Unit} \\ \hline
%    Resistance           & $R$    & \si\ohm \\
%    Resistivity          & $\rho$ & \si{\ohm\metre} \\
%    Length of conductor  & $L$    & \si\metre \\
%    Cross-sectional area & $A(x)$ & \si{\metre\squared}
%  \end{tabular}
%\end{center}



\begin{tabular}{c|c|c}
  \rowcolor{cyan}
           {\color{white}Gauge} & 
           {\color{white}Diameter} & 
           {\color{white}$R/L$} \\
           \rowcolor{cyan}
           & {\color{white}(\si{\milli\metre})} & 
                    {\color{white}(\SI{e-3}{\ohm\per\metre})}\\ \hline
                    0  & \num{9.35} & \num{0.31} \\
                    10 & \num{2.59} & \num{2.20} \\
                    14 & \num{1.63} & \num{8.54} \\
                    18 & \num{1.02} & \num{21.90} \\
                    22 & \num{0.64} & \num{51.70}
\end{tabular}

\begin{tabular}{c|c}
  \rowcolor{cyan}
           {\color{white} Material} & 
           {\color{white} Resistivity $\rho$ (\si{\ohm.\metre})}\\ \hline
           silver    & \num{1.6e-8} \\
           copper    & \num{1.7e-8} \\
           aluminum  & \num{2.7e-8} \\
           tungsten  & \num{5.6e-8} \\
           Nichrome  & \num{100e-8} \\
           carbon    & \num{3500e-8}\\
           germanium & \num{.46} \\
           glass     & \num{e10} to \num{e14}
\end{tabular}
  





\section{Ohm's Law}
The electric potential difference $V$ across a ``load'' (resistor) equals the
product of the current $I$ through the load and the resistance $R$ of the
load.
\begin{equation}
  \boxed{
    V=IR
  }
\end{equation}
%\begin{center}
%  \begin{tabular}{l|c|c}
%    \rowcolor{pink}
%    \textbf{Quantity} & \textbf{Symbol} & \textbf{SI Unit} \\ \hline
%    Potential difference & $V$    & \si\volt \\
%    Current              & $I$    & \si\ampere \\
%    Resistance           & $R$    & \si\ohm
%  \end{tabular}
%\end{center}
A resistor is considered ``ohmic'' if it obeys Ohm's law. Note that Ohm's law
is not a fundamental law in physics.



%\section{Ohmic Devices}
In an ohmic device such as a resistor, the relationship between voltage
across the device and the current through the device is linear, i.e.:
\begin{center}
  \begin{tikzpicture}[scale=.9]
    \draw[function] (0,0)--(2.5,2.5)
    node[midway,below right]{$\text{slope}=R$};
    \draw[axes] (0,0)--(3,0) node[right]{$I$};
    \draw[axes] (0,0)--(0,3) node[right]{$V$};
  \end{tikzpicture}
  \hspace{.15in}
  \begin{tikzpicture}[scale=.9]
    \draw[function] (0,0)--(2.5,2.5)
    node[pos=.65,below right]{$\text{slope}=\dfrac1R$};
    \draw[axes] (0,0)--(3,0) node[right]{$V$};
    \draw[axes] (0,0)--(0,3) node[right]{$I$};
  \end{tikzpicture}
\end{center}




%\section{Non-Ohmic Devices}
Loads that are \textbf{non-ohmic} do not obey Ohm's law: the relationship
between voltage and current is \emph{not} linear.
\begin{center}
  \begin{tikzpicture}[scale=.9]
    \draw[axes] (0,0)--(3,0) node[right]{$V$};
    \draw[axes] (0,0)--(0,3) node[right]{$I$};
    \draw[function,domain=0:2.45,smooth] plot(\x,.0008*\x^9);
    \end{tikzpicture}
\end{center}
For example, diodes (e.g.\ your TV's LED screen) are non-ohmic.




\section{Power Dissipated by a Resistor}
Power is the rate at which work $W$ is done, and from electrostatics, the
change in electric potential energy $\Delta E_q$ is proportional to the
amount of charge $q$ and the voltage $V$. This gives a very simple expression
for power through a resistor:  
\begin{equation}
  P=\diff Wt=\diff {E_q}t=\diff{(qV)}t=\left(\diff qt\right)V
  \;\rightarrow\;\boxed{P=IV}
\end{equation}
Combining Ohm's law with the above equation gives two additional expressions
for power through a resistor:
\begin{equation}
  \boxed{
    P=\frac{V^2}R
  }\quad\quad
  \boxed{
    P=I^2R
  }
\end{equation}



\section{Kirchhoff's Current Law}
The electric current that flows into any junction in an electric circuit must
be equal to the current which flows out.

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[thick]
%    \fill circle (.07);
%    \draw (0,2)--(0,-1.5) to[R] (0,-3);
%    \begin{scope}[rotate around={-60:(0,0)}]
%      \draw (0,0)--(0,-1.5);
%      \draw (0,-1.5) to[C,l=$C$,rotate around={-60:(0,-1.5)}] (0,-3);
%    \end{scope}
%    \draw[rotate=60] (0,0)--(0,-2);
    \draw (.2,1) to[short,o-] (1.5,1)--(3.5,1)--(3.5,-.2);
    \draw (1.5,1)--(1.5,-.2);
    \draw (2.5,1)--(2.5,-.2);
    \foreach \x in {1,...,3}
    \draw[vector,red] (\x+.5,.9)--(\x+.5,.1) node[midway,left]{$I_{\x}$};
    \draw[vector,red] (.4,1)--(1.3,1) node[midway,above]{$I$};
  \end{tikzpicture}
\end{figure}


In the example on the left, with $I$ going into the junction, and $I_1$,
$I_2$ and $I_4$ coming out, the current law says that
\begin{equation}
  I=I_1+I_2+I_3
\end{equation}
Basically, it means that there cannot be any accumulation of charges
anywhere in the circuit. The law is a consequence of conservation of charge.
  




\section{Kirchhoff's Voltage Law}
The voltage changes around any closed loop in the circuit must sum to zero,
no matter what path you take through an electric circuit. This also applies
regardless of what circuit devices are along the path. In the example shown
in Fig.~\ref{fig:kvl1}, a resistor $R$, capacitor $C$ and an inductor $L$ are
in parallel with a battery of emf $\mathcal E$.

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[scale=1.5,american voltages,thick]
    \draw (0,0) to[battery1,l=$\mathcal E$] (0,1.5)
    --(1.5,1.5) to[R=$R$] (1.5,0)--(0,0);
    \draw (1.5,1.5)--(3,1.5) to[C,l=$C$] (3,0)--(1.5,0);
    \draw (3,1.5)--(4.5,1.5) to[L,l=$L$] (4.5,0)--(3,0);
    \node[below] at (0,0) {\small$A$};
    \node[above] at (0,1.5) {\small$B$};
    \node[above] at (1.5,1.5) {\small$C$};
    \node[above] at (3,1.5) {\small$D$};
    \node[above] at (4.5,1.5) {\small$E$};
    \node[below] at (4.5,0) {\small$F$};
    \node[below] at (3,0) {\small$G$};
    \node[below] at (1.5,0) {\small$H$};
  \end{tikzpicture}
  \caption{A circuit with multiple circuit devices}
  \label{fig:kvl1}
\end{figure}

Assume that the current flows clockwise and we draw a clockwise loop, we get
\begin{equation}
  \mathcal E-V_R=0\quad\rightarrow\quad\mathcal E-IR=0
\end{equation}
  




\section{Resistors in Parallel}

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[thick]
    \draw (0,2) to[short,o-] (1,2) to[R=$R_1$] (1,0) to[short,-o] (0,0);
    \draw (1,2) to[short] (2.25,2) to[R=$R_2$] (2.25,0) to[short] (1,0);
    \draw (2.25,2) to[short] (3.5,2) to[R=$R_3\cdots$] (3.5,0)
    to[short] (2.25,0);
  \end{tikzpicture}
\end{figure}

The total current is the current through all the resistors, which can be
rewritten in terms of voltage and resistance using Ohm's law:
\begin{equation}
  I=I_1+I_2+I_3\cdots=\frac{V_1}{R_1}+\frac{V_2}{R_2}+\frac{V_3}{R_3}\cdots
\end{equation}    
Since $V_1=V_2=V_3=\cdots=V$ from the voltage law, we can re-write as
\begin{equation}
  I=\frac V{R_p}=V\left(\frac1{R_1}+\frac1{R_2}+\frac1{R_3}+\cdots\right)
\end{equation}
The inverse of the equivalent resistance for $N$ resistors connected in
parallel is the sum of the inverses of the individual resistances.
\begin{equation}
  \boxed{
    \frac1{R_p}=\sum_i^N\frac1{R_i}
  }
\end{equation}
%\begin{center}
%  \begin{tabular}{l|c|c}
%    \rowcolor{pink}
%    \textbf{Quantity} & \textbf{Symbol} & \textbf{SI Unit} \\ \hline
%    Equivalent resistance in parallel & $R_p$ & \si\ohm \\
%    Resistance of individual loads    & $R_i$ & \si\ohm
%  \end{tabular}
%\end{center}
As the number of resistors in parallel increase, the equivalent resistance
\emph{decreases} because this effectively increases the cross-sectional
area of the resistor.




\section{Resistors in Series}

\begin{figure}[ht]
  \centering
  \begin{tikzpicture}[thick]
    \draw (0,0) to[R=$R_1$,o-] (2,0) to[R=$R_2$] (4,0) to[R=$R_3$,-o] (6,0);
  \end{tikzpicture}
\end{figure}

The analysis for resistors in series is similar (but easier). From the current
law, the current through each resistor is the same:
\begin{equation}
  I_1=I_2=I_3=\cdots=I
\end{equation}
And the total voltage drop across all resistor is therefore:
\begin{equation}
  V=V_1+V_2+V_3+\cdots=I(R_1+R_2+R_3+\cdots)
\end{equation}
The equivalent resistance of loads is the sum of the resistances of the
individual loads.  
\begin{equation}
  \boxed{R_s=\sum_{i=1}^NR_i}
\end{equation}
%\begin{center}
%  \begin{tabular}{l|c|c}
%    \rowcolor{pink}
%    \textbf{Quantity} & \textbf{Symbol} & \textbf{SI Unit} \\ \hline
%    Equivalent resistance in series & $R_s$ & \si\ohm \\
%    Resistance of individual loads  & $R_i$ & \si\ohm
%  \end{tabular}
%\end{center}
Increasing the number of resistors in series \emph{increases} the equivalent
resistance because this effectively increases the length of the resistor.



%\section{Example Problem (Simple)}
%  A simple circuit analysis problem will involve one voltage source and
%  resistors connected, some in parallel, and some in series. Below is a typical
%  example:
%
%  \vspace{.2in}
%  
%    \begin{tikzpicture}[scale=.8,american voltages]
%      \draw(0,0) to[R=2<\ohm>](0,2) to[battery=100<\volt>] (0,4) to[short]
%      (1,4) to [R=10<\ohm>] (3,4)--(3,4.5) to[R=40<\ohm>] (5,4.5)
%      to[short] (5,4) to[short] (5.5,4)--(5.5,0)--(0,0);
%      \draw(3,4)--(3,3.5) to[R,l_=10<\ohm>] (5,3.5)--(5,4);
%      \draw[dashed] (-1.8,0.25) rectangle(0.6,3.75);
%    \end{tikzpicture}

%    Two \SI{10}{\ohm} resistors and a \SI{40}{\ohm} resistor are connected as
%    shown to a \SI{100}{\volt} emf source with internal resistance
%    \SI{2}{\ohm}. How much power is dissipated by the \SI{40}{\ohm} resistor?
%    \begin{enumerate}[(A)]
%    \item\SI{160}{\watt}
%    \item\SI{40}{\watt}
%    \item\SI{400}{\watt}
%    \item\SI{5}{\watt}
%    \item\SI{500}{\watt}
%    \end{enumerate}
%  
%



\section{Tips for Solving ``Simple'' Circuit Problems}
  \begin{enumerate}
  \item Identify groups of resistors that are in parallel or in series, and
    find their equivalent resistance.
  \item Gradually reduce the entire circuit to one voltage source and one
    resistor.
  \item Using Ohm's law, find the current out of the battery.
  \item Using Kirchhoff's laws, find the current through each of the resistors.
  \end{enumerate}





\section{Multi-loop Circuit}
Some of these problems require you to solve a system of linear equations.
The following is a simple example with two voltage sources:
\begin{center}
  \begin{tikzpicture}[american voltages,scale=1.3,thick]
    \draw (0,0) to[battery,l=$\mathcal E_1$] (0,2) to[R,l=$R_1$] (2,2)
    to[R,l=$R_3$] (2,0)--(0,0);
    \draw (2,0)--(4,0) to[battery,l_=$\mathcal E_2$] (4,2)
    to[R,l_=$R_2$] (2,2);

    \draw[vector,red] (2,.4)--(2,0) node[midway,right]{$I_1+I_2$};
    \draw[vector,red] (.6,.4) ..controls (0,2) and (2,2).. (1.6,.4)
    node[midway,below]{$I_1$};
    \draw[vector,red] (3.4,.4) ..controls (4,2) and (2,2).. (2.4,.4)
    node[midway,below]{$I_2$};
  \end{tikzpicture}
\end{center}
In this case, we have to draw two loops of current.




%\section{A More Difficult Example}
%\begin{center}
%  \begin{tikzpicture}[american voltages,thick]
%    \draw (0,0) to[battery, l=$\mathcal E_1$] (0,2) to[R=$R_1$] (2,2)
%    to[R=$R_3$] (2,0)--(0,0);
%    \draw (2,0)--(4,0) to[battery,l_=$\mathcal E_2$] (4,2) to[R,l_=$R_2$]
%    (2,2);
%    \begin{scope}[vector,red]
%      \draw (2,.4)--(2,0) node[midway,right]{$I_1+I_2$};
%      \draw (.6,.4) ..controls (0,2) and (2,2).. (1.6,.4)
%      node[midway,below]{$I_1$};
%      \draw (3.4,0.4) ..controls (4,2) and (2,2).. (2.4,.4)
%      node[midway,below]{$I_2$};
%    \end{scope}
%  \end{tikzpicture}
%\end{center}
  
We split the circuit into two loops, and apply Kirchkoff's voltage in both:
\begin{align*}
  \mathcal E_1-I_1R_1-(I_1+I_2)R_3&=0\\
  \mathcal E_2-I_2R_2-(I_1+I_2)R_3&=0
\end{align*}
There are two equations and two unknowns ($I_1$ and $I_2$) which we can either
solve algebraically or numerically. %We can subtract
%  (2) from (1), then solve for $I_1$ and $I_2$:
%
%  \begin{equation}
%    I_1=\frac{V_1-I_2R_3}{R_1+R_3}
%    \quad\quad
%    I_2=
%    \frac{\left[V_2-\frac{(V_1-V_2)R_3}{R_1}\right]}
%         {\left[R_2+\frac{(R_1+R_2)R_3}{R_1}\right]}
%  }
(Try this at home as an exercise.)



%\section{As Difficult As It Gets}
%  
%    \begin{tikzpicture}[american voltages,scale=1.2,thick]
%      \draw (0,0)--(0,2) to[R=3<\ohm>] (0,4)
%      to[battery1,l=42<\volt>] (2,4) to[R=3<\ohm>] (4,4)--(4,2)
%      to[R=4<\ohm>](4,0)--(2,0) to[battery1,l=6<\volt>] (2,2)
%      to[R,l_=4<\ohm>] (0,2);
%      \draw (0,0) to[R,l_=6<\ohm>](2,0);
%      \draw (2,2) to[R,l_=6<\ohm>](4,2);
%      \uncover<2->{
%        \begin{scope}[vector,red]
%          \draw (.5,3.5) ..controls(4.75,4) and (4.75,2).. (.5,2.5)
%          node[midway,left]{$I_1$};
%          \draw (.6,.4) ..controls (0,2) and (2,2).. (1.4,.4)
%          node[midway,below]{$I_3$};
%          \draw (2.6,.4) ..controls (2,2) and (4,2).. (3.4,.4)
%          node[midway,below]{$I_2$};
%        \end{scope}
%      }
%    \end{tikzpicture}

%    \begin{itemize}
%    \item To solve this problem, we define a few ``loops'' around the circuit:
%      one on top, one on bottom left, and one on bottom right.
%    \item<2-> Apply the voltage law in the loops. For example, in the
%      lower left:
%
%      \begin{equation}
%        4(I_1-I_3)-6-6I_3=0
%      }      
%    \item<2-> Solve the linear system to find the current. If the current that
%      you worked out is negative, it means that you have the direction wrong.
%    \end{itemize}

